Fail Faster

### 2025-03-17 paper/paper.tex
halfway trhough 3.1

Because the inlined versions of the generator are identical to the un-inlined
versions except for mechanical, semantics-preserving transformations,
this performance difference is attributable solely to
the different machine code generated by the compiler.
Indeed, compilers of effectful and strict functional languages (including JIT
compilers, in the case of Scala 3)
use heuristics to determine if and when to perform this particular kind of
simplification~\footnote{As we discuss in Section~\ref{asdf} \jwc{fixme}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.}.  Even in cases as simple as
Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions.

% Modern compilers use heuristics to decide when to specialize and inline code \cn, and these heuristics cannot guarantee that generator abstractions are zero-cost.
% These heuristics are also necessarily conservative about inlining and specializing recursive functions, which all generators of recursive data types must be.
% The

@@

% \jwc{The overhead of the abstraction is 2x. The reality is actually worse: actual base-quickcheck includes even more indirection in its type.}

% \jwc{The native code OCaml compiler, even with -O3, fails to specialize this code and eliminate the overhead of this abstraction --- inlining all of the funciton definitions and then performing beta-reductions speeds up sampling by a factor of 2.}

% \jwc{Same story in scala. 
% Approximately 2x difference for the same generator (though an order of magnitude slower overall than bq)... (scala uses a slightly different generator type, but the principles are the same)
% This doesn't even account for more inlining we could do to eliminate the boxing/unboxing of results. (Scalacheck generators can fail, though this adds significant overhead)
% }

% [info] GenBm.generateComplexBase     avgt    3  458.325 ± 67.214  ns/op
% [info] GenBm.generateComplexInlined  avgt    3  266.035 ±  8.559  ns/op

% \jwc{
% 
% }

\tr{Each of these consist of an explanation of the problem and pseudocode outlining the solution.}

\jwc{
  \begin{itemize}
    \item In cases where the compiler cannot statically eliminate them, running a monadic bind allocates a short-lived closure: we allocate a closure for the continuation, and then immediately jump into it.
    \item Each closure allocation is relatively cheap, but doing lots of allocation in a generation hot loop adds up fast.
    \item Similar things have been noticed about monadic parsing DSLs in the past.
  \end{itemize}
}

>>>>

Because the inlined versions of the generator are identical to the un-inlined
versions except for mechanical, semantics-preserving transformations,
this performance difference is attributable solely to
the different machine code generated by the compiler.
Indeed, compilers of effectful and strict functional languages (including JIT
compilers, in the case of Scala 3)
use heuristics to determine if and when to perform this particular kind of
simplification~\footnote{As we discuss in Section~\ref{asdf} \jwc{fixme}, purity
means that Haskell is a slightly different story. GHC can and often does
transformations of this form.}.  Even in cases as simple as
Figure~\ref{fig:overhead-explanation-code}, the indirection of \texttt{return}
and \texttt{bind} causes these heuristics to not fire.  Moreover, the story is
even worse for recursive generators, as the heuristics are necessarily even more
conservative for optimizing recursive functions. \jwc{unclear if we need this sentence, it's sorta obvious}

@@

While neither of these microbenchmarks exhibit this behavior, more complex
generators suffer further performance penalties due to \emph{closure
allocation}. In cases where the compiler cannot statically eliminate it \jwc{``it'' = ``the allocation''},
running a monadic bind allocates a short-lived closure: we allocate a closure
for the continuation, and then immediately jump into it.
In strict functional languages like OCaml and Scala,
Each individual closure allocation is relatively cheap, but doing lots of allocation in a generation
loop is very costly because each allocation brings us closer to the next costly GC pause \cn.
